---
title: 'Lab 10: Gibbs processes'
---

```{r, include = FALSE}
# This option turns evaluation of R code off in this document. Remove it
# if you use it as a template for solving the lab exercise.
knitr::opts_chunk$set(eval = TRUE)
```

This session is concerned with Gibbs models for point patterns with interpoint interaction.  
The lecturer's R script is [available here](https://raw.githubusercontent.com/spatstat/SSAI2017/master/Scripts/script10.R) (right click and save).

```{r}
library(spatstat)
```

### Exercise 1

In this question we fit a Strauss point process model to
the `swedishpines` data.

1.  We need a guess at the interaction distance $R$. Compute and
    plot the $L$-function of the dataset and choose the value $r$
    which maximises the discrepancy $\lvert L(r)-r \rvert$.

    We plot the above function which we want to maximize.

    ```{r}
    plot(Lest(swedishpines), abs(iso - r) ~ r, main = "")
    ```

    As seen from the plot, the maximum lies around $r = 6.5$ by eye.
    We find the optimum explicitly like follows:

    ```{r}
    discrep <- function(r) {
      return(abs(as.function(Lest(swedishpines))(r) - r))
    }
    res <- optimise(discrep, interval = c(0.1, 20), maximum = TRUE)
    print(res)
    R <- res$maximum
    ```
    
    This corresponds nicely with the plot.
    
2.  Fit the stationary Strauss model with the chosen interaction
    distance using
    ```{r, eval = FALSE}
    ppm(swedishpines ~ 1, Strauss(R))
    ```
    where `R` is your chosen value.

3.  Interpret the printout: how strong is the interaction?

4.  Plot the fitted pairwise interaction function using
    `plot(fitin(fit))`.

    As we have assigned `R`, we simply write:
    
    ```{r}
    fit <- ppm(swedishpines ~ 1, Strauss(R))
    print(fit)
    ```

    As seen, the $\gamma = 0.14$ parameter is quite small. Thus there seems to be a
    strong negative association between points within distance R of each other.
    A $\gamma$ of $0$ implies the hard core process whereas $\gamma = 1$ implies
    the Poisson process and thus CSR.

    The pairwise interaction function become:
    
    ```{r}
    plot(fitin(fit))
    ```

### Exercise 2

In Question 1 we guesstimated the Strauss interaction distance 
parameter. Alternatively this parameter could be estimated by 
profile pseudolikelihood.

1.  Look again at the plot of the $L$-function of
    `swedishpines` and determine a plausible range of
    possible values for the interaction distance.
    
    ```{r}
    plot(Lest(swedishpines), main = "")
    ```

    A conservative range of plausible interaction distances seems to be 3 to 15 meters.

2.  Generate a sequence of values equally spaced across this range,
    for example, if your range of plausible values was
    $[0.05, 0.3]$, then type
    ```{r, eval = FALSE}
    rvals <- seq(0.05, 0.3, by=0.01)
    ```

    We generate the numbers between 3 and 12.
    ```{r}
    rvals <- seq(3, 12, by = 0.1)
    ```

3.  Construct a data frame, with one column named `r`
    (matching the argument name of `Strauss`), containing
    these values. For example
    ```{r, eval = FALSE}
    D <- data.frame(r = rvals)
    ```

    OK, 
    
    ```{r}
    D <- data.frame(r = rvals)
    ```


4.  Execute
    ```{r, eval = FALSE}
    fitp <- profilepl(D, Strauss, swedishpines ~ 1)
    ```
    to find the maximum profile pseudolikelihood fit.

    OK, let's execute it: 
    
    ```{r}
    fitp <- profilepl(D, Strauss, swedishpines ~ 1)
    ```

5.  Print and plot the object `fitp`.

    ```{r}
    print(fitp)
    plot(fitp)
    ```

6.  Compare the computed estimate of interaction distance $r$ with
    your guesstimate. Compare the corresponding estimates of the
    Strauss interaction parameter $\gamma$.
    
    ```{r}
    (Ropt <- reach(as.ppm(fitp)))
    ```

    The $r = `r Ropt`$ is not totally inconsistent with the previous estimate of $7$.

7.  Extract the fitted Gibbs point process model from the object
    `fitp` as 
    ```{r, eval = FALSE}
    bestfit <- as.ppm(fitp)
    ```

    OK, let's do that:

    ```{r}
    bestfit <- as.ppm(fitp)
    ```

### Exercise 3

For the Strauss model fitted in Question 1,

1.  Generate and plot a simulated realisation of the fitted model
    using `simulate`.

    ```{r}
    s <- simulate(fit, drop = TRUE)
    plot(s, main = "")
    ```

2.  Plot the $L$-function of the data along with the global
    simulation envelopes from 19 realisations of the fitted model.

    ```{r}
    plot(envelope(fit, Lest, global = TRUE, nsim = 19, nsim2 = 100), main = "")
    ```



### Exercise 4

1.  Read the help file for `Geyer`.

    See `help(Geyer)`

2.  Fit a stationary Geyer saturation process to `swedishpines`, with the 
    same interaction distance as for the Strauss model computed in Question 2,
    and trying different values of the saturation parameter `sat = 1, 2, 3` say.

    ```{r}
    ppm(swedishpines ~ 1, Geyer(r = Ropt, sat = 1))
    ppm(swedishpines ~ 1, Geyer(r = Ropt, sat = 2))
    ppm(swedishpines ~ 1, Geyer(r = Ropt, sat = 3))
    ```

3.  Fit the same model with the addition of a log-quadratic trend.

    ```{r}
    gfit <- ppm(swedishpines ~ polynom(x, y, 2), Geyer(r = Ropt, sat = 3))
    ```

4.  Plot the fitted trend and conditional intensity.

    Here we use the log scale to be able to see the discs in the conditional intensity.
    ```{r}
    par(mfrow=c(1,2))
    plot(gfit, log = TRUE, pause = FALSE)
    ```


### Exercise 5

Modify question 1 by using the Huang-Ogata approximate
maximum likelihood algorithm (`method="ho"`) instead of maximum
pseudolikelihood (the default, `method="mpl"`).

```{r}
fit.mpl <- ppm(swedishpines ~ 1, Strauss(R), method = "mpl")
fit.ho  <- ppm(swedishpines ~ 1, Strauss(R), method = "ho")
print(fit.ho)
print(fit.mpl)
```

The fits are very similar.

### Exercise 6

Repeat Question 2 for the inhomogeneous Strauss process with
log-quadratic trend. The corresponding call to `profilepl` is
```{r, eval = FALSE}
fitp <- profilepl(D, Strauss, swedishpines ~ polynom(x,y,2))
```

```{r}
fitp2 <- profilepl(D, Strauss, swedishpines ~ polynom(x,y,2))
print(fitp)
print(fitp2)
```

### Exercise 7

Repeat Question 3 for the inhomogeneous Strauss process with
log-quadratic trend, using the inhomogeneous $L$-function `Linhom`
in place of the usual $L$-function.

```{r}
fit2 <- as.ppm(fitp2)
plot(envelope(fit2, Linhom, global = TRUE, nsim = 19, nsim2 = 100), main = "")
```
