---
title: 'Lab 4: Fitting Poisson models'
---

```{r, include = FALSE}
# This option turns evaluation of R code off in this document. Remove it
# if you use it as a template for solving the lab exercise.
knitr::opts_chunk$set(eval = TRUE)
```

This session is concerned with Poisson point process models.  
The lecturer's R script is [available here](https://raw.githubusercontent.com/spatstat/SSAI2017/master/Scripts/script04.R) (right click and save).

```{r message=FALSE}
library(spatstat)
```

### Exercise 1

The command `rpoispp(100)` generates realisations of the
Poisson process with intensity $\lambda = 100$ in the unit square.

1.  Repeat the command `plot(rpoispp(100))` several times
    to build your intuition about the appearance of a completely
    random pattern of points.

    Let's plot it three times:
    
    ```{r, results='hide'}
    replicate(3, plot(rpoispp(lambda = 100), main = ""))
    ``` 

    As can be seen, the points (unsurprisingly) are much more random that want one 
    might think. "Randomly" drawing points on a piece of paper one would usually
    draw a point pattern that is more regular (i.e. the points are repulsive).

2.  Try the same thing with intensity $\lambda = 1.5$.

    For brevity we only do it once here:
    
    ```{r, results='hide'}
    plot(rpoispp(lambda = 1.5), main = "")
    ``` 

    Here we expect 1.5 points in the plot each time.

### Exercise 2

Returning to the Japanese Pines data,

1.  Fit the uniform Poisson point process model to the Japanese
    Pines data
    ```{r, eval = FALSE}
    ppm(japanesepines~1)
    ```

    We fit the Poisson process model with the given command and print the output:
    
    ```{r}
    m.jp <- ppm(japanesepines ~ 1)
    print(m.jp)
    ```

2.  Read off the fitted intensity. Check that this is the correct
    value of the maximum likelihood estimate of the intensity.

    We extract the coeficient with the `coef` function, and compare to the
    straightforward estimate obtained by `intensity``:

    ```{r}
    unname(exp(coef(m.jp)))
    intensity(japanesepines)
    ```

    As seen, they agree exactly.


### Exercise 3

The `japanesepines` dataset is believed to exhibit
spatial inhomogeneity.

1.  Plot a kernel smoothed intensity estimate.

    Plot the kernel smoothed intensity estimate selecting the bandwidth with `bw.scott`:
    
    ```{r}
    jp.dens <- density(japanesepines, sigma = bw.scott)
    plot(jp.dens)
    plot(japanesepines, col = "white", cex = .4, pch = 16, add = TRUE)
    ```

2.  Fit the Poisson point process models with loglinear intensity
    (trend formula `~x+y`) and log-quadratic intensity (trend
    formula `~polynom(x,y,2)`) to the Japanese Pines data.

    We fit the two models with `ppm`:
    
    ```{r}
    jp.m <- ppm(japanesepines ~ x + y)
    jp.m2 <- ppm(japanesepines ~ polynom(x, y, 2) )
    ```

3.  extract the fitted coefficients for these models using
    `coef`.
        
    ```{r}
    coef(jp.m)
    coef(jp.m2)
    ```
    
4.  Plot the fitted model intensity (using `plot(predict(fit))`)

    ```{r}
    par(mar=rep(0,4))
    plot(predict(jp.m), main = "")
    plot(predict(jp.m, se=TRUE)$se, main = "")
    
    plot(predict(jp.m2), main = "")
    plot(predict(jp.m2, se=TRUE)$se, main = "")
    ```

5.  perform the Likelihood Ratio Test for the null hypothesis of a
    loglinear intensity against the alternative of a log-quadratic
    intensity, using `anova`.

    ```{r}
    anova(jp.m, jp.m2)
    ```

6.  Generate 10 simulated realisations of the fitted log-quadratic
    model, and plot them, using `plot(simulate(fit, nsim=10))` where 
    `fit` is the fitted model.

    ```{r}
    par(mar=rep(0.5,4))
    plot(simulate(jp.m2, nsim=10), main = "")
    ```


### Exercise 4

The `update` command can be used to re-fit a point
process model using a different model formula.

1.  Type the following commands and interpret the results:
    ```{r}
    fit0 <- ppm(japanesepines ~ 1)
    fit1 <- update(fit0, . ~ x)
    fit1
    fit2 <- update(fit1, . ~ . + y)
    fit2
    ```

    OK, let's do that:
    
    ```{r}
    fit0 <- ppm(japanesepines ~ 1)
    fit1 <- update(fit0, . ~ x)
    fit1
    fit2 <- update(fit1, . ~ . + y)
    fit2
    ```

2.  Now type `step(fit2)` and interpret the results.

    The backwards selection is done with the code:
    
    ```{r}
    step(fit2)
    ```
    
    First, given two models the preferred model is the one with the minimum AIC 
    value. In step 1, the removal of x results in the least AIC and is hence 
    deleted. In step 2, removing y results in a lower AIC than not deleing anything 
    and is thus deleted. This results in the constant model.
    
### Exercise 5

The `bei` dataset gives the locations of trees in a survey area
with additional covariate information in a list `bei.extra`.

1.  Fit a Poisson point process model to the data which assumes
    that the intensity is a loglinear function of terrain slope
    and elevation (hint: use `data = bei.extra` in `ppm`).

    We fit the log-linear intensity model with the following:
    
    ```{r}
    bei.m <- ppm(bei ~ elev + grad, data = bei.extra)
    ```

2.  Read off the fitted coefficients and write down the fitted
    intensity function.

    The coefficents are extraced with `coef`:
    
    ```{r}
    coef(bei.m)
    ```

    Hence the model is $log\lambda(u) = -8.55 + 0.02\cdot E(u) + 5.84 G(u)$ where
    $E(u)$ and $G(u)$ is the elevation and gradient, respectively, at $u$.

3.  Plot the fitted intensity as a colour image.

    ```{r}
    plot(predict(bei.m), main = "")
    plot(bei, cex = 0.3, pch = 16, cols = "white", add = TRUE)
    ```

4.  extract the estimated variance-covariance matrix of the
    coefficient estimates, using `vcov`.

    We call `vcov` on the fitted model object:
    ```{r}
    vcov(bei.m)
    ```

5.  Compute and plot the standard error of the intensity estimate
    (see `help(predict.ppm)`).

    From the documentation the argument `se` will trigger the computation of the 
    standard errors. These are then plotted in the standard manner.
    
    ```{r}
    std.err <- predict(bei.m, se = TRUE)$se
    plot(std.err, main = "")
    ```


### Exercise 6

Fit Poisson point process models to the Japanese Pines data, with
the following trend formulas. Read off an expression for the fitted
intensity function in each case.

|Trend formula     |Fitted intensity function                   |
|:-----------------|:-------------------------------------------|
|`~1`              | $\log\lambda(u) = 4.17$                    |
|`~x`              | $\log\lambda(u) = 4.28 - 0.23x$            |
|`~sin(x)`         | $\log\lambda(u) = 4.29 - 0.26\sin(x)$      |
|`~x+y`            | $\log\lambda(u) = 4.07 - 0.23x + 0.42y$    | 
|`~polynom(x,y,2)` | $\log\lambda(u) = 4.06 + 1.14x - 1.56y - 0.75x^2 - 1.20xy + 2.51y^2$ |
|`~factor(x < 0.4)`| $\log\lambda(u) = 4.10 + 0.16\cdot I(x < 0.4)$ |

(Here, $I(\cdot)$ denote the indicator function.)

The fitted intensity functions have been written into the table based on the follwing model fits:

```{r}
coef(ppm1 <- ppm(japanesepines ~ 1)) 
coef(ppm2 <- ppm(japanesepines ~ x))
coef(ppm3 <- ppm(japanesepines ~ sin(x)))
coef(ppm4 <- ppm(japanesepines ~ x + y))
coef(ppm5 <- ppm(japanesepines ~ polynom(x, y, 2)))
coef(ppm6 <- ppm(japanesepines ~ factor(x < 0.4)))
```


### Exercise 7

Make image plots of the fitted intensities for the inhomogeneous
models above.

Again, we use `plot(predict())`:

```{r}
plot(predict(ppm1), main = "")
plot(predict(ppm2), main = "")
plot(predict(ppm3), main = "")
plot(predict(ppm4), main = "")
plot(predict(ppm5), main = "")
plot(predict(ppm6), main = "")
```

