---
title: 'Lab 7: Envelopes and Monte Carlo tests'
---

```{r, include = FALSE}
# This option turns evaluation of R code off in this document. Remove it
# if you use it as a template for solving the lab exercise.
knitr::opts_chunk$set(eval = TRUE)
```

This session is concerned with evelopes of summary statistics and Monte Carlo tests.  
The lecturer's R script is [available here](https://raw.githubusercontent.com/spatstat/SSAI2017/master/Scripts/script07.R) (right click and save).

```{r}
library(spatstat)
```

### Exercise 1

For the `swedishpines` data:

1.  Plot the $K$ function along with pointwise envelopes from 39
    simulations of CSR:
    ```{r, eval = FALSE}
    plot(envelope(swedishpines, Kest, nsim=39))
    ```

    OK,
    
    ```{r}
    plot(envelope(swedishpines, Kest, nsim=39), main = "")
    ```

2.  Plot the $L$ function along with pointwise envelopes from 39
    simulations of CSR.
    
    Like above now with `Lest`:
    
    ```{r}
    plot(envelope(swedishpines, Lest, nsim=39), main = "")
    ```

3.  Plot the $L$ function along with **simultaneous**
    envelopes from **19** simulations of CSR, using
    `ginterval=c(0,5)`.
    
    ```{r}
    plot(envelope(swedishpines, Lest, nsim = 19, global = TRUE, ginterval=c(0,5)), main = "")
    ```
    
4.  Plot the $L$ function for along with
    **simultaneous** envelopes from **99**
    simulations of CSR using `ginterval=c(0,5)`. What
    is the significance level of the associated test?

    ```{r}
    plot(envelope(swedishpines, Lest, nsim = 99, global = TRUE, ginterval=c(0,5)), main = "")
    ```

    Which yields an 1% significance test.

### Exercise 2

To understand the difficulties with the $K$-function when the point
pattern is not spatially homogeneous, try the following experiment
(like in the previous lab session).

1.  Generate a simulated realisation of an inhomogeneous Poisson
    process, e.g.
    ```{r, eval = FALSE}
    X <- rpoispp(function(x,y){ 200 * exp(-3 * x) })
    ```

    OK,
    
    ```{r}
    X <- rpoispp(function(x,y){ 200 * exp(-3 * x) })
    ```

2.  Compute simulation envelopes (of your favorite type) of the $K$-
    or $L$-function under CSR. They may well indicate significant
    departure from CSR.

    There indeed often seems to be a departure from CSR:
    
    ```{r}
    par(mfrow=c(1,2))
    plot(envelope(X, Kest, global = TRUE))
    plot(envelope(X, Lest, global = TRUE))
    ```

3.  Fit a Poisson point process model to the `japanesepines`
    data with log-quadratic trend (formula `~polynom(x,y,2)`). Plot the
    $L$-function of the data along with simultaneous envelopes from 99
    simulations **of the fitted model**.
    
    This can be done by the code:
    
    ```{r}
    fit <- ppm(japanesepines ~ polynom(x, y, 2))
    plot(envelope(fit, Linhom, global = TRUE), main = "")
    ```
