---
title: 'Lab 3: Intensity dependent on covariate'
---

```{r, include = FALSE}
# This option turns evaluation of R code off in this document. Remove it
# if you use it as a template for solving the lab exercise.
knitr::opts_chunk$set(eval = TRUE)
```

This session covers tools for investigating intensity depending on a covariate.  
The lecturer's R script is [available here](https://raw.githubusercontent.com/spatstat/SSAI2017/master/Scripts/script03.R) (right click and save).

```{r, message=FALSE}
library(spatstat)
```

### Exercise 1

The `bei` dataset gives the locations of trees in a survey area with additional
covariate information in a list `bei.extra`.

1.  Assign the elevation covariate to a variable `elev` by typing
    ```{r, eval = FALSE}
    elev <- bei.extra$elev
    ```

    OK, lets do that:
    
    ```{r}
    elev <- bei.extra$elev
    ```

2.  Plot the trees on top of an image of the elevation covariate.
    
    ```{r}
    plot(elev, main = "")
    plot(bei, add = TRUE, cex = 0.3, pch = 16, cols = "white")
    ```

3.  Cut the study region into 4 areas according to the value of the 
    terrain elevation, and make a texture plot of the result.
        
    ```{r}
    Z <- cut(elev, 4, labels=c("Low", "Med-Low", "Med-High", "High"))
    textureplot(Z, main = "")
    ```

4.	Convert the image from above to a tesselation, count the number 
    of points in each region using `quadratcount`, and plot the
    quadrat counts.
    
    ```{r}
    Y <- tess(image = Z)
    qc <- quadratcount(bei, tess = Y)
    ```

5.  Estimate the intensity in each of the four regions.
    
    ```{r}
    intensity(qc)
    ```


### Exercise 2

Assume that the intensity of trees is a function $\lambda(u) = \rho(e(u))$
where $e(u)$ is the terrain elevation at location u.

1.  Compute a nonparametric estimate of the function $\rho$ and plot it by
    ```{r, eval = FALSE}
    rh <- rhohat(bei, elev)
    plot(rh)
    ```

    Repeating the R code:
    
    ```{r}
    rh <- rhohat(bei, elev)
    plot(rh)
    ```

2.	Compute the predicted intensity based on this estimate of $\rho$.
    
    ```{r}
    prh <- predict(rh)
    plot(prh, main = "")
    plot(bei, add = TRUE, cols = "white", cex = .2, pch = 16)
    ```

3.  Compute a non-parametric estimate by kernel smoothing and
  	compare with the predicted intensity above.

    The kernel density estimate of the points is computed and plotted with the
    following code:
    
    ```{r}
    dbei <- density(bei, sigma = bw.scott)
    plot(dbei, main = "")
    plot(bei, add = TRUE, cols = "white", cex = .2, pch = 16)
    ```
    
    Which seems to be quite different form the predicted intentisty.

4.  Bonus info: To plot the two intensity estimates next to each
  	other you collect the estimates as a spatial object list
  	(`solist`) and plot the result (the estimates are called
  	`pred` and `ker` below):
    ```{r, eval = FALSE}
    l <- solist(pred, ker)
    plot(l, equal.ribbon = TRUE, main = "", 
         main.panel = c("rhohat prediction", "kernel smoothing"))
    ```

    ```{r}
    l <- solist(prh, dbei)
    plot(l, equal.ribbon = TRUE, main = "",
         main.panel = c("rhohat prediction", "kernel smoothing"))
    ```


### Exercise 3

1.  Continuing with the dataset `bei` conduct both Berman's Z1 and Z2 tests for
    dependence on `elev`, and plot the results.

    The tests are done straightforwardly with `berman.test`:
    
    ```{r}
    Z1 <- berman.test(bei, elev)
    print(Z1)
    plot(Z1)
    
    Z2 <- berman.test(bei, elev, which = "Z2")
    print(Z2)
    plot(Z2)
    ```
