---
title: 'Lab 8: Spacing and distances'
---

```{r, include = FALSE}
# This option turns evaluation of R code off in this document. Remove it
# if you use it as a template for solving the lab exercise.
knitr::opts_chunk$set(eval = TRUE)
```

This session is concerned with summary statistics for
spacings and interpoint distances.  
The lecturer's R script is [available here](https://raw.githubusercontent.com/spatstat/SSAI2017/master/Scripts/script08.R) (right click and save).

```{r}
library(spatstat)
```

### Exercise 1

For the `swedishpines` data:

1.  Calculate the estimate of the nearest neighbour distance
    distribution function $G$ using `Gest`.

    ```{r}
    G <- Gest(swedishpines)
    ```

2.  Plot the estimate of $G(r)$ against $r$

    ```{r}
    plot(G, cbind(km, rs, han) ~ r, main = "Nearest neighbor distance distribution")
    ```

3.  Plot the estimate of $G(r)$ against the theoretical (Poisson) value
    $G_{\mbox{pois}}(r) = 1 - \exp(-\lambda \pi r^2)$.

    E.g.
    
    ```{r}
    plot(G, . ~ theo, main = "Nearest neighbor distribution")
    ```

4.  Define Fisher’s variance-stabilising transformation for c.d.f.’s
    by
    ```{r, eval = FALSE}
    Phi <- function(x) asin(sqrt(x))
    ```
    Plot the $G$ function using the formula `Phi(.) ~ Phi(theo)`
    and interpret it.
    
    ```{r}
    Phi <- function(x) asin(sqrt(x))
    plot(G, Phi(.) ~ Phi(theo), main = "Nearest neighbor distribution")
    ```
    
    The transformation has made the deviations from CSR in the central part of the
    curve smaller. We need envelopes to say anything about significance.

### Exercise 2

For the `swedishpines` data:

1.  Calculate the estimate of the nearest neighbour distance
    distribution function $F$ using `Fest`.

    ```{r}
    Fhat <- Fest(swedishpines)
    ```

2.  Plot the estimate of $F(r)$ against $r$

    ```{r}
    plot(Fhat, main = "Empty Space function")
    ```

3.  Plot the estimate of $F(r)$ against the theoretical (Poisson) value
    $F_{\mbox{pois}}(r) = 1 - \exp(-\lambda \pi r^2)$.

    ```{r}
    plot(Fhat, . ~ theo, main = "")
    ```

4.  Define Fisher’s variance-stabilising transformation for c.d.f.’s
    by
    ```{r, eval = FALSE}
    Phi <- function(x) asin(sqrt(x))
    ```
    Plot the $F$ function using the formula `Phi(.) ~ Phi(theo)`
    and interpret it.

    ```{r}
    Phi <- function(x) asin(sqrt(x))
    plot(Fhat, Phi(.) ~ Phi(theo), main = "")
    ```
    
    The transformation has changed the picture much. It looks like deviation from CSR,
    but again, without envelopes it's hard to draw conclusions.
    