---
title: 'Lab 5: Marked point patterns'
---

```{r, include = FALSE}
# This option turns evaluation of R code off in this document. Remove it
# if you use it as a template for solving the lab exercise.
knitr::opts_chunk$set(eval = TRUE)
```

This session covers modelling of marked point patterns using marked
point processes.  
The lecturer's R script is [available here](https://raw.githubusercontent.com/spatstat/SSAI2017/master/Scripts/script05.R) (right click and save).

```{r message=FALSE}
library(spatstat)
```

### Exercise 1

The dataset `spruces` contains the Norwegian Spruces
dataset giving the locations of trees and their diameters at breast
height.

1.  Read the help file for the data;

2.  access the dataset and plot it;

    ```{r}
    plot(spruces)
    ```

3.  re-plot the data so that the tree diameters are displayed at a
    physical scale that is 10 times the physical scale of the
    location coordinates.

    ```{r}
    plot(spruces, markscale = 10)
    ```

4.  use `Smooth` (notice the upper case S) to compute and plot a
    pixel image of the local average tree diameter;

    ```{r}
    plot(Smooth(spruces, sigma = 8))
    ```

5.  trees are normally classified as ‘adult’ when their diameter
    exceeds 30 centimetres. Use the `cut` command to
    classify each tree as adult or juvenile, and produce a multitype
    point pattern in which the trees are marked as adult or
    juvenile. Plot this pattern, and plot the adults and juveniles
    separately.
    
    ```{r}
    cutspruces <- cut(spruces, breaks = c(0,0.3,0.4), labels = c("juvenile", "adult"))
    plot(cutspruces, main = "")
    plot(split(cutspruces), main = "")
    ```


### Exercise 2

The file `anthills.txt` is available in the Data directory on 
[github](https://github.com/spatstat/SSAI2017) and downloadable by 
[this direct link](https://raw.githubusercontent.com/spatstat/SSAI2017/master/Data/anthills.txt) (right click and save).

It records the locations of anthills recorded in a
1200x1500 metre study region in northern Australia.
Coordinates are given in metres, along with a letter code recording
the ecological ‘status’ of each anthill.

1.  read the data into `R` as a data frame, using the `R` function
    `read.table`. (Since the input file has a header line, you
    will need to use the argument `header=TRUE` when you call
    `read.table`.)

    ```{r}
    dat <- read.table(file = "../Data/anthills.txt", header = TRUE)
    ```

2.  check the data for any peculiarities.

3.  create a point pattern `hills` containing these data.
    Ensure that the marks are a factor, and that the unit of length
    is given its correct name.

    ```{r}
    hills <- with(dat, ppp(x, y, xrange = c(0, 1200), yrange = c(0, 1500),
                           marks = status, units=c("metre", "metres")))
    ```

4.  plot the data.

    ```{r}
    plot(hills, main = "")
    ```
    

### Exercise 3

The dataset `hamster` is a multitype pattern representing
the locations of cells of two types, *dividing* and *pyknotic*.

1.  plot the data;

    ```{r}
    plot(hamster)
    ```

2.  plot the patterns of pyknotic and dividing cells separately;

    ```{r}
    plot(split(hamster), main = "")
    ```

3.  plot kernel estimates of the intensity functions of pyknotic and
    dividing cells separately;

    ```{r}
    plot(density(split(hamster)), main = "")
    ```

4.  use `relrisk` to perform cross-validated bandwidth
    selection and computation of the relative intensity of pyknotic
    cells.

    ```{r}
    plot(relrisk(hamster, hmax = 1, relative = TRUE, control = "dividing"))
    ```

### Exercise 4

The command `rmpoispp` generates simulated realisations
of a *multitype* Poisson process. The first argument
`lambda` specifies the intensity function
$\lambda(x,y,m)$ which gives the intensity of points at location
(x,y). It may be given in several forms.

1.  If `lambda` is a single number, it specifies the
    intensity of points *of each type*. Try computing, inspecting
    and plotting the result of
    ```{r, eval=FALSE}
    rmpoispp(21, win=square(1), types=c("yes", "no"))
    ```
    What is the expected total number of points generated by this
    command?
    
    ```{r}
    X <- rmpoispp(21, win=square(1), types=c("yes", "no"))
    summary(X)
    plot(X, main = "")
    ```

    We expect $2\times21=42$ points on average.

2.  If `lambda` is a vector of numbers, the vector
    entries specify the intensities for each type of point. Try
    ```{r, eval=FALSE}
    rmpoispp(c(20,40,20), types=letters[1:3])
    ```
    What is the expected total number of points generated by this
    command?
    
    ```{r, eval=FALSE}
    rmpoispp(c(20,40,20), types=letters[1:3])
    ```

    We expect $20+40+20=80$ points on average in this case.

3.  If `lambda` is a function with arguments
    `x,y,m` then this is interpreted as the intensity
    function $\lambda(x,y,m)$. Try
    ```{r}
    fun <- function(x,y,m) { 40 * (x+y) }
    X <- rmpoispp(fun, types=letters[24:26])
    fun2 <- function(x,y,m) { ifelse(m == "yes", 100 * x, 50 * (1-x)) }
    X2 <- rmpoispp(fun2, types=c("yes", "no"))
    ```
    What is the expected total number of points in `X`?

    We expect $3\cdot\int_{[0,1]}\int_{[0,1]} 40 (x+y) d\!x\!d\!y = 120$
    points on average in this case.

### Exercise 5

Take the Harkness-Isham ants’ nests data `ants`

1.  use `summary` to estimate the average intensities of
    the points of each type.

    ```{r}
    summary(ants)
    lam <- intensity(ants)
    ```

2.  Generate and plot a realisation of a marked Poisson process in
    the same window as the data, with the same possible types of
    points, with uniform intensities for each type, given by the
    intensities estimated from the data.

    ```{r}
    rmpoispp(lam, win = Window(ants), types = names(lam))
    ```
    
    Or the even shorter
    
    ```{r}
    rmpoispp(lam, win = Window(ants))
    ```

3.  Repeat the simulation several times. Do the simulations look
    like the data?

    Make 11 simulations:
    ```{r}
    simants <- rmpoispp(lam, win = Window(ants), types = names(lam), nsim = 11)
    ```
    
    Add original data as list item number 12:
    ```{r}
    simants[[12]] <- ants
    ```

    Plot them all:
    ```{r}
    plot(simants, legend = FALSE, main = "")
    ```


### Exercise 6

Here we will fit multitype Poisson point process models to the
Harkness-Isham ants’ nests data `ants`.

1.  Fit the model `ppm(ants ~ marks)` and interpret the result.
    Compare the result with `summary(ants)` and explain
    the similarities.

    ```{r}
    fit1 <- ppm(ants ~ marks)
    ```
    
    This is a Poisson model with a separate constant intensity for each mark. The
    fitted intensities are:
    ```{r}
    exp(coef(fit1)[1])
    exp(coef(fit1)[1] + coef(fit1)[2])
    ```

    This agrees perfectly with the output of `summary(ants)`:
    ```{r}
    summary(ants)
    ```

2.  Fit the model `ppm(ants ~ marks + x)` and write down an
    expression for the fitted intensity function.

    ```{r}
    fit2 <- ppm(ants ~ marks + x)
    (co <- coef(fit2))
    ```
    
    Intensity for the reference type (Cataglyphis):
    
    $$\lambda( (x,y) ) = \exp(`r co[["(Intercept)"]]` + `r co[["x"]]` \cdot x)$$

    Intensity for the other type (Messor):

    $$\lambda( (x,y) ) = \exp(`r co[["(Intercept)"]]` + `r co[["marksMessor"]]` + `r co[["x"]]` \cdot x)$$

3.  Fit the model `ppm(ants ~ marks * x)` and write down an
    expression for the fitted intensity function.

    ```{r}
    fit3 <- ppm(ants ~ marks * x)
    (co <- coef(fit3))
    ```
    
    Intensity for the reference type (Cataglyphis):
    
    $$\lambda( (x,y) ) = \exp(`r co[["(Intercept)"]]` + `r co[["x"]]` \cdot x)$$
    
    Intensity for the other type (Messor):

    $$\lambda( (x,y) ) = \exp(`r co[["(Intercept)"]]` + `r co[["marksMessor"]]` + (`r co[["x"]]` + `r co[["marksMessor"]]`) \cdot x)$$

4.  Compute the fitted intensities of the three models fitted above 
    using `predict` and plot the results.

    ```{r}
    pred <- c(predict(fit1), predict(fit2), predict(fit3))
    plot(as.solist(pred), ncols = 2, main = "")
    ```

5.  Explain the difference between the models fitted by
    `ppm(ants ~ marks + x)` and `ppm(ants ~ marks * x)` .

    For the additive model the effect of the `x` coordinate is the same for both types
    of ants, while the effect of `x` differs in the multiplicative model.

### Exercise 7

The study region for the ants’ nests data `ants` is
divided into areas of ‘scrub’ and ‘field’. We want to fit a Poisson
model with different intensities in the field and scrub areas.

The coordinates of two points on the boundary line between field and
scrub are given in `ants.extra$fieldscrub`. First construct a function that
determines which side of the line we are on:
```{r}
fs <- function(x,y) {
  ends <- ants.extra$fieldscrub
  angle <- atan(diff(ends$y)/diff(ends$x))
  normal <- angle + pi/2
  project <- (x - ends$x[1]) * cos(normal) + (y - ends$y[1]) * sin(normal)
  factor(ifelse(project > 0, "scrub", "field"))
}
```

Now fit the models:
```{r, eval=FALSE}
ppm(ants ~ marks + side, data = list(side=fs))
ppm(ants ~ marks * side, data = list(side=fs))
```
and interpret the results.

```{r}
fit1 <- ppm(ants ~ marks + side, data = list(side=fs))
fit2 <- ppm(ants ~ marks * side, data = list(side=fs))
```

In the first model the fitted intensity is lower in the scrub than in the field
(but this effect is not significant).

In the second model the fitted intensity of Cataglyphis is lower in the scrub
than the intensity of Cataglyphis in the field, where as it is the other way
around for Messor. When we allow for the different effect between ant types the
scrub/field covariate is significant.
